
code, the actual ffi etc

documentation how to run the program

document the API, idris maybe haskell as well?

\section{Original Project Plan}\label{ch:origplan}
\begin{itemize}

    \item week 1 --- 29 Oct - 04 Nov,
          week 2 --- 05 Nov - 11 Nov, \\
          week 3 --- 12 Nov - 18 Nov ,
          week 4 --- 19 Nov - 25 Nov \\
          Consult more related work.
          Background reading.
          Unikernels and Library OS. \\
          FFIs.
          Experiment with Mirage.
          Experiment with Idris.
          Experiment with Ocaml

    \item week 5 --- 26 Nov - 02 Dec,
          week 6 --- 03 Dec - 09 Dec \\
          Configure development environment,
          Requirements gathering,
          Build Specification
    \item week 7 --- 10 Dec - 16 Dec \\
          Experiment with Mirage
    \item week 8 --- 17 Dec - 23 Dec,
          week 9 --- 24 Dec - 30 Dec \\
          Start implementing small bits
          of the ffi, essentially,
          work towards the first objective
    \item week 10 --- 31 Dec - 06 Jan \\ Fix Idris ergonomic issues
    \item \textbf{week 11 --- 07 Jan - 13 Jan ---
              Friday 11, 09:00
              Progress Presentation} \\ Preliminary progress presentation
          reparation, plan for the second semester
    \item week 12 --- 14 Jan - 20 Jan,
          week 13 --- 21 Jan - 27 Jan \\
          Implement basic deliverable
          and start writing
          the report
    \item week 14 --- 28 Jan - 03 Feb,
          week 15 --- 04 Feb - 10 Feb \\ Implement the intermediate
          deliverable and report Testing
    \item week 16 --- 11 Feb - 17 Feb,
          week 17 --- 18 Feb - 24 Feb,
          week 18 --- 25 Feb - 03 Mar \\ Attempting the advanced
          deliverable, Report Testing, \textbf{Mirage}
    \item week 19 --- 04 Mar - 10 Mar,
          week 20 --- 11 Mar - 17 Mar \\
          \textbf{week 21 --- 18 Mar - 24 Mar ---
              Need to be ready by now} \\
          Validation,
          Verification,
          Proofread \\
          Refactor,
          Finish Writing up,
          Prepare Demo
    \item \textbf{week 22 --- 25 Mar - 27 Mar ---
              Submission Monday 25, 12:00
              Demo is on Wednesday 27} \\ Hooray!

\end{itemize}

\section{Docs}
%todo

how to run how to do clone rep etc

\subsection{Findlib sorcery}

Put that Docker image see if it works

documentation of the IO API