\documentclass[a4paper, 12pt, oneside]{report}
\usepackage{setspace}
\onehalfspacing
\usepackage[outputdir=_build]{minted}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage[hidelinks]{hyperref}
% \usepackage{courier}
% \usepackage{fancyhdr} todo want header?
% \pagestyle{fancy}
\usepackage{listings}
\lstset{basicstyle=\footnotesize\ttfamily, captionpos=b}
\usepackage{natbib} 
\bibliographystyle{plainnat}
\usepackage[nottoc]{tocbibind}

% \title{Real World Idris}
% \author{Ioan Luca \\ 201638554 \\ 
% 	\small University of Strathclyde} % todo title
% \date{\today}

\begin{document}

\begin{titlepage}
	\centering
	{\scshape\LARGE Submitted for the Degree of B.Sc. in Computer Science, 2018/19 \par}
	\vspace{1cm}
	{\huge\bfseries Real World Idris\par}
	\vspace{1.5cm}
	{\scshape\Large Ioan Luca \par}
	\vspace{2cm}
	{\Large\itshape Except where explicitly stated all the work in this report, including appendices, is my own and was carried out during my final year. It has not been submitted for assessment in any other context.\par}
	\vfill
	201638554 

	\vfill

% Bottom of the page
	{\large \today\par}
\end{titlepage}

% \maketitle 

\pagenumbering{roman}

% \listoffigures
% \listoftables

\begin{abstract} 
Idris is a general purpose, dependently typed programming language that has an
advanced type system which encodes complex properties about programs into
types, which means that the compiler checks the programs for correctness
before they run.

Idris is at the forefront of a new generation of programming languages
that support “lightweight verification” --- specifications are defined
early during development, enabling writing maintainable code that is
guaranteed to work.
This has the potential to drastically reduce the costs of reliable software by
integrating verification into development, instead of considering it a separate
concern.

``Real World Idris'' aims to upgrade the Idris ecosystem
with a Foreign Function Interface (FFI) to OCaml.
The idea is to enable the two-way interoperability between these two
languages.

This would essentially allow Idris programs to access the mature OCaml
library ecosystem, which would ultimately make Idris more suitable
for practical programming.
\end{abstract}

\chapter*{Acknowledgements} 
% \addcontentsline{toc}{chapter}{Acknowledgements}
First and foremost, I have to thank my research supervisor,  
Dr.~Robert Atkey.
Without his assistance and dedicated 
involvement in every step throughout the process, 
this work would have never been accomplished. 
I would like to thank you very much for your 
support and understanding over the past year and for
introducing me to the wonderful world of Programming Language Research.
 
\tableofcontents

\pagenumbering{arabic}

\chapter{Introduction}\label{ch:intro}
\input{chapters/introduction}

\chapter{Related Work}\label{ch:related}
\input{chapters/related}

ocaml still unpopular check index \autoref{ch:intro}

\chapter{Problem Description and Specification}\label{ch:spec}
\input{chapters/spec}

\chapter{System Design}\label{ch:design}
\input{chapters/design}

\chapter{Detailed Design and Implementation}\label{implementation}
\input{chapters/implementation}

\chapter{Verification and Validation}\label{verification}
\input{chapters/verification}

\chapter{Results and Evaluation}\label{evaluation}
\input{chapters/evaluation}

\chapter{Summary and Conclusions}\label{conclusion}
\input{chapters/conclusion}


\nocite{*}
\bibliography{bibliography}

\appendix
\chapter{Appendix Title}
\input{appendices/appendix}

\end{document}
